Fail Faster

### 2025-03-11 paper/paper.tex
staging is hard

Muli-stage programming (also known more simply as staging) is \jwc{finish}

\jwc{A good quick presentation of staged metaprogramming is here, in section 6.
https://www.cl.cam.ac.uk/~jdy22/papers/a-typed-algebraic-approach-to-parsing.pdf
}

\jwc{Staging has a long history as a technique for eliminating the runtime overhead of abstractions,
specializing programs at an earlier stage to run more efficiently at a later stage.}

\jwc{For the purposes of this paper, we will only use two stages: compile time, and run time.}

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, splice, stage distinction.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
    \item We can convert one way, but not the other.
    \item Functions \texttt{'a code -> 'b code} ``fuse''.
    \begin{itemize}
      \item Consider writing \texttt{even . succ}. This
      \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
      \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
      a dynamic function value eliminates the function abstraction.
    \end{itemize}
    \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
    abstractions the library introduces are fully eliminated at compile time.
  \end{itemize}
}

@@

\jwc{
  \begin{itemize}
    \item Scala. Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS (in Scala 2, and Macros in Scala 3). Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
    Al of this should still work in scala. Could easily be incorporated into ScalaCheck, with minor modification: ScalaCheck uses a state monad to thread around the seed, instead of a stateful one (like BQ), or a splittable one (like Haskell). So you have to adapt to that. But same diff.
    \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
    GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
    the highest-performance generators.
    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
  \end{itemize}
}

>>>>

\jwc{
  \begin{itemize}
    \item In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
with each stage producing the code to be run in the next.
\item This is an old idea, with roots going back to quasiquotation in LISP, and picking up steam again in the 1990s with MetaML \cn{}.
\item For the purposes of this paper, we will only consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
which produces more code, which is then compiled and run at run time.
\item Staging has many applications, but chief among them is its use for \emph{optimization}. We can write programs such that during the compile time stage, they are partially evaluated to eliminate
abstraction overhead. 
\item Many languages have some degree of staging functionality, either provided as a library \cn or build directly into the language's implementation \cn.
\item For this paper, we use MetaOCaml \cn for OCaml staging, but all of the staging concepts are portable to any other language with multi-stage programming functionality.
  \end{itemize}
}

\jwc{This section is going to need work.}

MetaOCaml's staging functionality is exposed through a type \texttt{'a code}. A
value of type \texttt{t code} at compile time is a (potentially open) OCaml term
of type \texttt{t}.

Values of \texttt{code} type are introduced by \emph{quotes}, written \texttt{.<$\ldots$>.}. Brackets
delay execution of a program until run time. For example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.
Note that this is not the same as \texttt{.< 6 >.}. Because brackets delay computation, the code is not \emph{executed}
until the next stage (run time).

Values of type \texttt{code} can be combined together using \emph{escape}, written \texttt{.\~{}($\ldots$)}.
Escape lets you take a value of type \texttt{code}, and ``splice'' it directly into a quote.
For example, this program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to \texttt{.<(1 * 5) + (1 * 5)>.}.

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, escape or ``splice'', stage distinction.
    \item MetaOCaml ensures correct scoping and hygene by preventing alpha-collision when you splice open terms.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
    \item We can convert one way, but not the other.
    \item Functions \texttt{'a code -> 'b code} ``fuse''.
    \begin{itemize}
      \item Consider writing \texttt{even . succ}. This
      \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
      \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
      a dynamic function value eliminates the function abstraction.
    \end{itemize}
    \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
    abstractions the library introduces are fully eliminated at compile time.
  \end{itemize}
}

@@

\jwc{
  \begin{itemize}
    \item Scala. Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS (in Scala 2, and Macros in Scala 3). Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
    Al of this should still work in scala. Could easily be incorporated into ScalaCheck, with minor modification: ScalaCheck uses a state monad to thread around the seed, instead of a stateful one (like BQ), or a splittable one (like Haskell). So you have to adapt to that. But same diff.
    \item Python. Maybe could do it?? Hypothesis + MacroPy
    \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
    GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
    the highest-performance generators.
    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
  \end{itemize}
}
